Modular Model-Checking for Fault-Tolerant Distributed System Implementations

### 2016-12-02 doc/om-paper/paper.tex
Updates to section 1, including several filled in citations

%\documentclass{IEEEtran/IEEEtran}
\documentclass{llncs/llncs}

@@

\title{Modular Model-Checking for Fault-Tolerant Distributed System Implementations}
\author{Ben and Lee}

@@

The verification of this class of systems has usually been at the algorithmic level, eliding details about a concrete implementation, and historically, has relied on formal models verified by interactive theorem-proving~\cite{om-acl2-impl,Young97:IC,csl-93-2,pvs}. If formal verification is to be introduced into the workflow of system designers, though, we need more automated methods. (Mostly) automated proof techniques are required to reduce the need for specialized verification expertise. We also need programmatic verification of \emph{implementations}. System designers create software and hardware implementations to test, simulate, and deploy. Discrepancies between implementations and algorithmic models can arise if the latter is abstracted too much from the former, particularly if those abstractions are ad-hoc and system specific. Furthermore, as an implementations are tweaked and modified to explore the design space, it is easy for the formal models to become inconsistent.

There are at least two classes of abstractions that separate traditional models of fault-tolerant distributed algorithms from their implementations. One is to intertwine the environmental model with the system description. For example, the behavior of nodes is naturally specified as a transition system in which transitions are guarded by the node's fault state. But faults are part of the environment; an implementation does not use its own fault status to choose actions! Another class of abstractions is used to simplify models. For example, message passing might be abstracted with shared state, or a node's local behavior is elided and instead, the output is constrained by a \emph{specification} of the behavior.

In this paper, we develop a formal model to address these concerns. We combine ideas from the literature and folklore to build scalable and modular formal models of fault-tolerant distributed systems. We describe a generic formal model for fault-tolerant distributed systems that aims to reduce the need for ad-hoc verification abstractions and optimizations while maintaining scalability in Section~\ref{sec:model}. There are three aspects of the model: (1) \emph{calendar automata}, originally developed by Dutertre and Sorea~\cite{cal} for specifying real-time systems; (2) \emph{synchronous observers}, originally from the synchronous languages community~\cite{} and popularized by Rushby for model-checking~\cite{}; and (3) \emph{abstract state machines}, useful for abstracting counterexamples. The model is suitable for infinite-state model-checking.

To demonstrate the model in Section~\ref{sec:byz}, we present a formally-verified implementation of the Oral Messages Hybrid algorithm (OMH)~\cite{}, implementing the Boyer-Moore Fast Majority Vote algorithm for algorithm~\cite{}. The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm. In Section~\ref{sec:experimental}, we give performance metrics for the model. First, we benchmark the model, showing that despite the additional complexity, it scales farther than other model-checking verifications at the algorithmic level. The scalability is due principally to specifying invariants and using $k$-induction in SRI's SAL model-checker~\cite{}. Developing invariants requires some user guidance. A user should not have to discover all new invariants for new protocols or implementations. We show that most of the lemmas required are not protocol specific and are modular, concerning only specific aspects of the system. To demonstrate this, we modify the OHM implementation in various ways to show that only local modifications are required to the model and proofs.

@@

%% To address this problem, we envision building a domain-specific language (DSL) for specifying and exploring distributed fault-tolerant system implementations. We envision a DSL that generates both code for simulation, testing, and deployment, as well as formal models for verification. That way, the formal models verified and code executed stay in synch, and abstractions are systematically made. We say more about the DSL in Section~\ref{sec:conclusions}.

@@

\subsection{Calendar Automata}\label{sec:calendar}
\emph{Timed automata} are specialized formalisms for verifying real-time systems~\cite{}. Timed automata assume the existence of continuously-varying real-valued variables, known as \emph{real-time clocks} in a solver. Model-checkers for timed-automata are specialized for verifying real-time systems in which the only infinite-valued variables are the real-time clocks.

@@

Calendar automata avoid the problems of encoding timed automata. Rather than encoding ``how much time has passed since the last event'', it encodes ``how far into the future is the next scheduled event'', and a real-valued variable representing the current time is updated to the next event time. Doing so avoids Zeno updates.

@@

\item \label{cal:d} In all states, if the current time equals a timeout, then the only transition enabled is a calendar event update associated with the timeout: $\forall \sigma \in S$, $\exists (e_i, t_i) \in \sigma(cal)$ such that $\sigma(now) = t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma'(now) = \sigma(now)$, $\sigma'(c_j) = \sigma(c_j)$ for all $c_j \in \sigma(c)$ such that $c_j \neq c_i$, and $\sigma'(t_i) > \sigma(t_i)$.
\end{enumerate}

>>>>

% vim: wrap linebreak nolist textwidth=0 wrapmargin=0
%\documentclass{IEEEtran/IEEEtran}
\documentclass{llncs/llncs}

@@

\title{Modular Model-Checking for Fault-Tolerant Distributed System Implementations}
\author{Benjamin F Jones and Lee Pike}
\institute{Galois, Inc.}

@@

The verification of this class of systems has usually been at the algorithmic level, eliding details about a concrete implementation, and historically, has relied on formal models verified by interactive theorem-proving~\cite{om-acl2-impl,Young97:IC,csl-93-2,pvs}. If formal verification is to be introduced into the workflow of system designers, though, we need more automated methods. (Mostly) automated proof techniques are required to reduce the need for specialized verification expertise. We also need programmatic verification of \emph{implementations}. System designers create software and hardware implementations to test, simulate, and deploy. Discrepancies between implementations and algorithmic models can arise if the latter is abstracted too much from the former, particularly if those abstractions are ad-hoc and system specific \ben{for example?}. Furthermore, as implementations are modified to explore the design space, it is easy for the formal model and the implementation to become inconsistent.

There are at least two classes of abstractions that separate traditional models of fault-tolerant distributed algorithms from their implementations.  One is to intertwine the environmental model with the system description. For example, the behaviors of nodes are naturally specified as a transition system in which transitions are guarded by the node's fault state. But faults are part of the environment; an implementation does not use its own fault status to choose actions! Another class of abstractions is used to simplify models.  For example, message passing might be abstracted with shared state, or a node's local behavior is elided and instead, the output is constrained by a \emph{specification} of the behavior, e.g. pre- and post-conditions.

In this paper, we develop a formal model to address these concerns. We combine ideas from the literature and folklore to build scalable and modular formal models of fault-tolerant distributed systems. We describe a generic formal model for fault-tolerant distributed systems that aims to reduce the need for ad-hoc verification abstractions and optimizations while maintaining scalability in Section~\ref{sec:model}. There are three aspects of the model: (1) \emph{calendar automata}, originally developed by Dutertre and Sorea~\cite{cal} for specifying real-time systems; (2) \emph{synchronous observers}, originally from the synchronous languages community~\cite{Halbwachs-Lagnier:92,Halbwachs-Lagnier:94} and popularized by Rushby for model-checking~\cite{Rushby:SAS14}; and (3) \emph{abstract state machines}, useful for abstracting counterexamples. The model is suitable for infinite-state model-checking.

To demonstrate the modeling technique, in Section~\ref{sec:byz} we present a formally-verified implementation of the Oral Messages Hybrid algorithm (OMH)~\cite{Lincoln-Rushby}, implementing the Boyer-Moore Fast Majority Vote algorithm for algorithm~\cite{Boyer-Moore:91}. The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm. In Section~\ref{sec:experimental}, we give performance metrics for the model. First, we benchmark the model, showing that despite the additional complexity, it scales farther than other model-checking verifications at the algorithmic level. The scalability is due principally to specifying invariants and using $k$-induction in SRI's SAL model-checker~\cite{sal}. Developing invariants requires some user guidance. A user should not have to discover all new invariants for new protocols or implementations. We show that most of the lemmas required are not protocol specific and are modular, concerning only specific aspects of the system. To demonstrate this, we modify the OHM implementation in various ways to show that only local modifications are required to the model and proofs.

@@

%% To address this problem, we envision building a domain-specific language (DSL) for specifying and exploring distributed fault-tolerant system implementations. We envision a DSL that generates both code for simulation, testing, and deployment, as well as formal models for verification. That way, the formal models verified and code executed stay in sync, and abstractions are systematically made. We say more about the DSL in Section~\ref{sec:conclusions}.

@@

\subsection{Calendar Automata}\label{sec:calendar}
\emph{Timed automata} are specialized formalisms for verifying real-time systems~\cite{}. Timed automata assume the existence of continuously-varying real-valued variables, known as \emph{real-time clocks}. Model-checkers for timed-automata are specialized for verifying real-time systems in which the only infinite-valued variables are the real-time clocks.

@@

The calendar automaton avoids the problems of encoding timed automata. Rather than encoding ``how much time has passed since the last event'', it encodes ``how far into the future is the next scheduled event'', and a real-valued variable representing the current time is updated to the next event time. Doing so avoids Zeno updates.

@@

\item \label{cal:d} In all states, if the current time equals a timeout, then the only transitions enabled are calendar event updates associated with the timeout: $\forall \sigma \in S$, $\exists (e_i, t_i) \in \sigma(cal)$ such that $\sigma(now) = t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma'(now) = \sigma(now)$, $\sigma'(c_j) = \sigma(c_j)$ for all $c_j \in \sigma(c)$ such that $c_j \neq c_i$, and $\sigma'(t_i) > \sigma(t_i)$.
\end{enumerate}
